\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-24 chapters/bci.tex
Set manuscript draft

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like accoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.
The technologies above all measure the hemodynamic signal, which only indirectly
reflects brain activity.
Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, it is relatively
impractical due to the size of the equipment needed.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
neurology practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
offer great temporal resolution up to thousands of Hz.
As it is easily applicable on the outside of the scalp, many practical
consumer-grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and
research systems often use electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

@@

Paradigms with high active participation can be split up in \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user,
such as imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive
communication methods often limit the considered domain to a few movement
directions to control a cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded
like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
Imagined speech or movements can be performed in a multitude of ways, which
can each have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing
poor performance and giving rise to the concept of \ac{bci}
illiteracy~\cite{Allison2010}\footnote{
  \cite{Becker2022,Thompson2019}
}.
\todo{footnote on how this just means it needs to be more adapted,
inclusivity, harmful to progress, paraphrase abstracts of articles}

@@

\subsection{Stimulation paradigm}
One by one, visual elements on a computer screen are \emph{intensified} for a
short period of time by changing in color, brightness or size.
On each flash observed by the user, specific brain activity is elicited.
If one of these intensified elements is rare with respect to the others, i.e. it is the
\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
A stimulus can be rare either due to its inherent properties like color or
location
Crucially, this is also the case if it is `marked' as rare by the user, e.g.
by paying explicit attention to only a given stimulus and not the others or
counting how many times it flashes.
If these visual elements represent letters, and we know the timing of the
intensifications of each letter, we can now establish communication by detecting
for which letter an oddball response was present in the brain signal at its
time of intensification.

\Ac{itr} in the oddball paradigm is optimized by intensifying
multiple targets at once in a row-collumnn strategy and using a sequence of
selections, giving rise to the classic
P300 matrix speller of~\cite{Farwell1988}.
Other optimizations increase response \ac{snr}, like using flashing face
images as intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
stimuli~\cite{Treder2011};

Sometimes Flashing in various configurations, such as row-column or in groups,
other flashing patterns
faces, feature attention
visual hex-o-spell

comfortable, preferred over SSVEP, similar ITR is this true?
\todo{figure of P3 matrix}
\todo{cite P300 literature, first example and current sota}

@@

later compoentns P3 more attention/cognitive more centrally in the brain

Because of this, the oddball paradigm is often referred to as the P300
paradigm.
However, due to equal importance of other \ac{erp} components than the P3 in
decoding as we will show later, this could be considered a bit of a misnomer
and we will stick to the oddball paradigm naming.

mention VEPs more from visual system occipitally

P300 and other components, explain function (Luck)

\todo{figure of ERP components}

which components, at what amplitude and latency they are evoked varies per
person and depends on the type of intensification

\subsubsection{P1}
\subsubsection{N1}
\subsubsection{P2}
\subsubsection{N2}
\subsubsection{P3}
\subsubsection{Late negativity}

>>>>

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.
The technologies above all measure the hemodynamic signal, which only indirectly
reflects brain activity.
Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, it is relatively
impractical due to the size of the equipment needed.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
neurology practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
offer great temporal resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volt (V), often on the scale of microvolt ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical
consumer-grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and
research systems often use active electrodes with preamplifiers and
electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality, since they require a trained
professional to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

@@

Paradigms with high active participation can be split up in \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user,
such as imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive
communication methods often limit the considered domain to a few movement
directions to control a cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded
like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
Imagined speech or movements can be performed in a multitude of ways, which
can each have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing
poor performance and giving rise to the concept of \ac{bci}
illiteracy~\cite{Allison2010}\footnote{
  \cite{Becker2022,Thompson2019}
}.
\todo{footnote on how this just means it needs to be more adapted,
inclusivity, harmful to progress, paraphrase abstracts of articles}
\todo{user training more than with reactive, task instruction}

@@

\subsection{Stimulation paradigm}
One by one, visual elements on a computer screen are \emph{intensified} for a
short period of time by changing in color, brightness or size.
Intensifcations of different targets are usually 100-200ms apart and last for about
100ms~\cite{Sellers2006a}.
On each flash observed by the user, specific brain activity is elicited.
If one of these intensified elements is rare with respect to the others, i.e. it is the
\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
A stimulus can be rare either due to its inherent properties like color or
location
Crucially, this is also the case if it is `marked' as rare by the user, e.g.
by paying explicit attention to only a given stimulus and not the others or
counting how many times it flashes.
If these visual elements represent letters, and we know the timing of the
intensifications of each letter, we can now establish communication by detecting
for which letter an oddball response was present in the brain signal at its
time of intensification.

\Ac{itr} in the oddball paradigm is optimized by intensifying
multiple targets at once in a row-collumnn strategy and using a sequence of
selections, giving rise to the classic
P300 matrix speller of~\cite{Farwell1988}.
Other optimizations increase response \ac{snr}, like using flashing face
images as intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
stimuli~\cite{Treder2011}.
The number of targets in a visual oddball \ac{bci}is limited by the crowding
effect, which imposes a limit on how close targets can be to each other while
not distracting the user's attention~\cite{Sellers2006a,Li2010}.
This may be overcome by making hierarchical selections of stimuli representing
groups of selections~\cite{Treder2010}.

Together with \ac{ssvep}, oddball paradigms are most frequently used in visual
\ac{bci}.
There are however indications that users prefer oddball stimulation over
\ac{ssvep} which can cause eye strain and fatigue due to the continuous
oscillatory stimulation of all targets at once~\cite{Xu2021}.
Furthermore, \ac{ssvep} relies more on directing the gaze at the intended
target than oddball, detracting from the concept of control independent from
muscle movement.

\todo{four pane figure:
  photo of someone using a P300 bci
  P300 matrix interface
  simplified figure of ERP components
  full spatiotemporal ERP response
}

@@

The time-domain response ellicited immediately after intensification of a
stimulus is the visual \ac{erp}~\cite{Luck2014}.
An \ac{erp} is a waveform consisting of multiple components, of wich some are
modulated by whether the stimulus was an oddball or not.
Components are positive peaks and negative troughs in the \ac{erp} waveform.
They following a naming scheme according to their polarity (Negative or
Positive) and their latency (e.g. 3 or 300ms).
The component most prominently modulated by attention is called the P300 or P3
(positive deflection after $\pm$300ms) and oddball paradigms are therefore
frequently referred to as P300 paradigms.
However, due to the equally important contribution of other \ac{erp}
components than the P3 in decoding as we will show later, this could be
considered a bit of a misnomer and we will stick to the oddball paradigm naming.
P300 and other components, explain function (Luck)
Which components, at what amplitude and latency they are evoked varies per
person and depends on the exact details of the stimulation.

In general, earlier components, such as the \acp{vep}, are more tied to sensory
information processing and less modulated by attention, while later components
are more related to attention and cognitive processes.

\textcite{Luck2013}
\subsubsection{P1}
\subsubsection{N1}
\subsubsection{P2}
\subsubsection{N2}
\subsubsection{P3}
\subsubsection{Late negativity}

### 2024-09-24 chapters/conclusion.tex
BCI chapter
Add gaze disabilities
Add decoders
Literature wcble
Set manuscript draft

\chapter{Conclusions \& recommendations}
\section{General discussion}
\section{Limitations}
\section{Future directions}
\section{Working with patients}
\section{Conclusion}

\section{Obtained results}

\subsection{Regularized spatiotemporal beamforming}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (xDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher accuracy in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\cite{VanDenKerchove2022}.

@@

\subsection{Healthy control study}
The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
$26.38\pm3.15$ years.
This dataset gives us insight in the ERP dynamics in overt, covert and the
novel split VSA condition, and confirms our hypothesis that P3 jitter has a
significant impact on performance in covert and split VSA, as well as that the
amplitude effect observed for covert VSA also holds for split VSA.

@@

>>>>

\todo{https://sci-hub.ru/https://ieeexplore.ieee.org/abstract/document/1642754}
\todo{https://www.sciencedirect.com/science/article/abs/pii/B9780444639349000020}
\todo{BCI principles and practice chapter 19}

\section{General discussion}

\subsection{Obtained results}

\subsubsection{Regularized spatiotemporal beamforming}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (xDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher accuracy in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\cite{VanDenKerchove2022}.

@@

\subsubsection{Healthy control study}
The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
$26.38\pm3.15$ years.
This dataset gives us insight in the ERP dynamics in overt, covert and the
novel split VSA condition, and confirms our hypothesis that P3 jitter has a
significant impact on performance in covert and split VSA, as well as that the
amplitude effect observed for covert VSA also holds for split VSA.

@@

\section{Limitations}

\section{Current \& future work}
multicomponent alignment approach
hankel tensors to cope with jitter
select best classifier using eye tracking

\section{Recommendations}
\subsection{Working with patients}
start from patients
use all available signals
\subsection{Linear decoding methods}

\section{Conclusion}

### 2024-09-24 chapters/covert_align.tex
Literature wcble
Set manuscript draft

\section{Introduction}

@@

The participant was instructed to press the space bar when ready for a block
of stimulations.
Then, one target was indicated as the cue and the participant
was instructed to count the number of intensifications of the cued target
during the following block of stimulations.
After pressing the space bar again, a blue crosshair appeared, and the subject
was instructed to fixate their gaze on the blue crosshair for the duration of
the stimulation block (Figure~\ref{fig:layout_cross_overt} and Figure~\ref{fig:layout_cross_covert}).
The position of this crosshair determined the VSA condition for this trial:
overt VSA when the crosshair was at the same location as the cued target,
covert VSA when the crosshair appeared in the center of the screen, and split
VSA when the crosshair appeared on a different target than the cued one.
After pressing the space bar again and a delay of 5 seconds, the stimulation block
starts.
All targets were intensified for a duration of 100 ms, in pseudorandom order.
The inter-stimulus-interval (ISI), the time between the onsets of subsequent intensifications,
was variable and consisted of a fixed 300ms interval (of which 100ms with an intensified target onscreen)
with 200ms uniform jitter added, resulting in an ISI
between 200 and 400 ms.
Inter-stimulus intervals were jittered to counteract steady-state effects and residue in averaging. A
longer inter-stimulus interval will increase component amplitude and aid in counteracting temporal
autocorrelation for a higher statistical test precision.
In a block of stimulations, each target was intensified a pseudorandom number of times between 10 and 15.
This led to stimulation blocks with an average duration of 26.25 seconds. After a block of stimulations, an
input prompt appeared to enter the mentally counted number of intensifications.
After inputting this number, the subject was allowed to pause until pressing the space bar again.
In total, six blocks were presented for overt VSA, six blocks for covert VSA, 12 blocks for
split ($d=1$) VSA, 12 blocks for split ($d=2$) VSA and 6 blocks for split
($d=3$) VSA, covering all possible combinations of VSA conditions, cued targets and
crosshair locations.
The experiment started with five non-recorded practice stimulation blocks, one for
each of the 5 VSA conditions.
During these practice blocks, the participant received feedback about their gaze
position and counting accuracy.
Counting the instructions and the participant's response to the
input prompts, a block lasted about 30
seconds. In sum, the experiment featured approximately 45 minutes of
stimulation time.
After blocks 14 and 28, the participant was allowed to take a longer break.
Including these longer breaks, the experiment lasted approximately one hour.
\subsubsection{Riemannian Geometry}
We will compare WCBLE, CBLE and tLDA to Riemannian Geometry
approaches that rely on spatial covariance as features.
Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
performance~\cite{Lotte2018}.
We implemented two Riemannian Geometry pipelines.
The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
filters, projects these covariances to a tangent space and classifies the
result using $L2$-regularized logistic regression
(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
favorable performance in the presence of single-trial ERP latency jitter.
Shrunk spatial covariance matrices are estimated from epochs that are
augmented by concatenating the average target and average non-target ERP as
extra channels, projected to tangent space and classified using
$L2$-regularized logistic regression (ERPCov-TS-LR).

@@

>>>>

\section{Introduction}
\todo{mention cble and wcble in methods}

@@

The participant was instructed to press the space bar when ready for a block
of stimulations.
Then, one target was indicated as the cue and the participant
was instructed to count the number of intensifications of the cued target
during the following block of stimulations.
After pressing the space bar again, a blue crosshair appeared, and the subject
was instructed to fixate their gaze on the blue crosshair for the duration of
the stimulation block (Figure~\ref{fig:layout_cross_overt} and Figure~\ref{fig:layout_cross_covert}).
The position of this crosshair determined the VSA condition for this trial:
overt VSA when the crosshair was at the same location as the cued target,
covert VSA when the crosshair appeared in the center of the screen, and split
VSA when the crosshair appeared on a different target than the cued one.
After pressing the space bar again and a delay of 5 seconds, the stimulation block
starts.
All targets were intensified for a duration of 100 ms, in pseudorandom order.
The inter-stimulus-interval (ISI), the time between the onsets of subsequent intensifications,
was variable and consisted of a fixed 300ms interval (of which 100ms with an intensified target onscreen)
with 200ms uniform jitter added, resulting in an ISI
between 200 and 400 ms.
Inter-stimulus intervals were jittered to counteract steady-state effects and residue in averaging. A
longer inter-stimulus interval will increase component amplitude and aid in counteracting temporal
autocorrelation for a higher statistical test precision.
In a block of stimulations, each target was intensified a pseudorandom number of times between 10 and 15.
This led to stimulation blocks with an average duration of 26.25 seconds. After a block of stimulations, an
input prompt appeared to enter the mentally counted number of intensifications.
After inputting this number, the subject was allowed to pause until pressing the space bar again.
In total, six blocks were presented for overt VSA, six blocks for covert VSA, 12 blocks for
split ($d=1$) VSA, 12 blocks for split ($d=2$) VSA and 6 blocks for split
($d=3$) VSA, covering all possible combinations of VSA conditions, cued targets and
crosshair locations.
The experiment started with five non-recorded practice stimulation blocks, one for
each of the 5 VSA conditions.
During these practice blocks, the participant received feedback about their gaze
position and counting accuracy.
Counting the instructions and the participant's response to the
input prompts, a block lasted about 30
seconds. In sum, the experiment featured approximately 45 minutes of
stimulation time.
After blocks 14 and 28, the participant was allowed to take a longer break.
Including these longer breaks, the experiment lasted approximately one hour.

@@

\subsubsection{Decoders}
We will compare \ac{wcble} and \ac{cble} ad described in
Section~\ref{sec:wcble/methods}, with their base classifier and tLDA and with
Riemannian Geometry
approaches that rely on spatial covariance as features.
Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
performance~\cite{Lotte2018}.
We implemented two Riemannian Geometry pipelines.
The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
filters, projects these covariances to a tangent space and classifies the
result using $L2$-regularized logistic regression
(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
favorable performance in the presence of single-trial ERP latency jitter.
Shrunk spatial covariance matrices are estimated from epochs that are
augmented by concatenating the average target and average non-target ERP as
extra channels, projected to tangent space and classified using
$L2$-regularized logistic regression (ERPCov-TS-LR).
